\vspace{-1in} Demystifying the Computer Science \\PhD Admission in US Universities\\{\large A Guide for International Students

### 2023-11-06 demystify.tex
updates

\emph{Funding}:  In Europe, stipend (salary) comes from the university or from gov't research organizations and often have a fixed duration (e.g., 3 years).  In the US, stipend (e.g., RA) comes directly from your adviser (no fixed duration).  There are also fewer TA opportunities in the European universities compared to the US.

\emph{Faculty position after PhD}: In other countries, PhD graduates typically apply for research positions at research labs, i.e., postdocs, and then consider faculty position. In the US, PhD graduates interested in academia can apply directly for faculty position.  This is why US PhD programs are longer and more focused on research.

\emph{Work-life balance}: PhD students are often overworked in the US (seriously!) compared to other countries, e.g., in Europe.  This is partly due to the longer PhD program and that US PhD students are often paid through TA, which requires them to work for the dept. In contrast, PhD students in other countries are often paid through fellowships, which do not require them to do TA work.

@@

Note that in some universities the formal title Dr. Lastname is preferred. You just need to observe and follow the conventions at your particular institution. Additionally, be aware that not all faculty members may hold a Ph.D., in which case using Prof. Lastname is a suitable alternative.

\paragraph{Referring to professors you know} Because you are already familiar with these individuals, you can just informally use their firstname as mentioned above (or Dr./Prof., if you want to be formal). You can also include their institution if it makes it more precise.  For example, I can say:  \emph{"I did my postdoc with Jeff Foster at Univ. of Maryland"}.

Do not include ranking (e.g., Assistant, Associate, Scientist, ...) when referring to someone. I see many international students include a lengthy title of people they know, e.g., \emph{I am advised by Asst. Prof. X, and I also collaborate with Distinguished Scientist Y}.  This is \emph{not necessary} and makes it look like you're trying to show off your connections. These nuances represent some cultural differences that you may encounter and will gradually adapt to. More on cultural differences in \S\ref{sec:cultural}.

@@

\subsection{Will I be miserable during my a PhD?}\label{sec:happy}
Many posts from VietPhD, a Facebook group for Vietnamese PhD students, are about how PhD students are mistreated, stressed, and miserable. Certainly there are bad relationships with professors, co-authors, and lab mates, and toxic working environment, feeling left out or discriminated (e.g., because you're an international student), and it is good to be aware of those.  However, in reality there are many good mentors, supportive lab mates and department, happy students, etc.  So don't let social media make you feel pessimistic and deter your quest to advance knowledge.

@@

Here is the typically steps: (i) a faculty suspecting a cheating case will report it to the Office of Academic Integrity (OAI) at the university;  (ii)  OAI will take over and investigate the case; and finally (iii) OAI will make the final decision.  It is important to note that after receiving the report from your prof., OAI \emph{completely} takes over and makes decision.  This means begging your professor will not help because they simply are no longer involved in the case.

@@

There's a misconception that in the US it's all business, with professors as bosses who pay students for their work, and you doing something nice implying you expect something in return. However, the reality is quite the opposite. While people can be straightforward and direct, they are also friendly and informal. You can call your professor by their first name (\S\ref{sec:address}), respectfully disagree and argue with that (and can actually gain respect doing so), seek their help (even on personal matters), come to their houses for parties or gathering, and give them small thoughtful gifts that they put on their desks.  Many students and professors maintain lifelong relationships, staying in touch through cards, emails, and calls, even after their academic journey ends. These connections are genuine and meaningful.

@@

\paragraph{About Me} I am an assistant professor in the CS dept at George Mason University (GMU). Prior to GMU, I was at the University of Nebraska-Lincoln (UNL). I have served in the PhD admission process at GMU and UNL for many years.  Currently I am also the program director of the Masters program in Software Engineering at GMU (thus also have some experience with the MS admission process--which is quite different than PhD). I got my undergraduate and MS degrees from Penn State. I completed my PhD at the University of New Mexico-Albuquerque and a postdoc at the University of Maryland-College Park.

While I grew up in the US, I am familiar with international students, e.g., many of my students are international students.  I have many collaborators from various countries and have been a mentor for many students from Vietnam.  I also have many colleague and friends who were international students and have heard their stories.  I hope to capture their challenges and experiences in this document so that it can be useful for prospective international students.

>>>>

\emph{Funding}:  In many countries stipend comes from the university or from gov't research organizations and often have a fixed duration (e.g., 3 years).  In the US, stipend (e.g., RA) comes directly from your adviser (no fixed duration).  There are also fewer TA opportunities in the European universities compared to the US.

\emph{Faculty position after PhD}: In other countries, PhD graduates who are interested academia typically apply for research positions at research labs, i.e., postdocs, and then consider faculty position. In the US, PhD graduates often apply directly for faculty position (postdoc for US graduates is no longer a popular option as it was before).  This is why US PhD programs are longer and more focused on research.

\emph{Work-life balance}: I believe that PhD students are often overworked in the US compared to other countries, e.g., in Europe.  This is partly due to the longer PhD program and that US PhD students are often paid through TA, which requires them to do TA in addition to their own research. In contrast, PhD students in other countries are often paid through fellowships, which do not require them to do TA work.

@@

Note that in some universities the formal title \emph{Dr.} Lastname is preferred. You just need to observe and follow the conventions at your particular institution. Additionally, be aware that not all faculty members may hold a Ph.D., in which case using Prof. Lastname is a suitable alternative.

\paragraph{Referring to professors you know} Because you are already familiar with these individuals, you can just informally use their names if they are OK with it as mentioned above (or Dr./Prof., if you want to be formal). You can also include their institution if it makes it more precise.  For example, I can say:  \emph{"I did my postdoc with Jeff Foster at Univ. of Maryland"}.

Do not include ranking (e.g., Assistant, Associate, Scientist, ...) when referring to someone. I see many students include a lengthy title of people they know, e.g., \emph{I am advised by Asst. Prof. X, and I also collaborate with Distinguished Scientist Y}.  This is \emph{not necessary} and makes it look like you're trying to show off your connections. These nuances represent some cultural differences that you may encounter and will gradually adapt to. More on cultural differences in \S\ref{sec:cultural}.

@@

\subsection{Will I be miserable during my a PhD?}\label{sec:happy}
Many posts from VietPhD, a Facebook group for Vietnamese PhD students, are about how PhD students are mistreated, stressed, and miserable. Certainly issues including bad relationships with professors, conflicts with co-authors and lab mates, and toxic working environment, feeling left out or discriminated (e.g., because you're an international student) \emph{do} exist, and it is good to be aware of those.  However, in reality there are many good mentors, supportive lab mates and department, happy students, etc.  So don't let social media make you feel pessimistic and deter your quest to advance knowledge.

@@

Here is the typically steps: (i) a faculty suspecting a cheating case will report it to the Office of Academic Integrity (OAI) at the university---the report often has supporting evidence and suggested penalty (e.g., receive a failing grade);  (ii)  OAI will take over and investigate the case; and (iii) OAI will make the final decision.  It is important to note that after receiving the report from your prof., OAI \emph{completely} takes over and makes decision.  This means begging your professor will not help because they simply are no longer involved in the case.

@@

There's a misconception that in the US it's all business, with professors as bosses who pay students for their work, and you doing something nice implying you expect something in return. However, the reality is quite the opposite. While people can be straightforward and direct, they are also friendly and informal. You can call your professor by their first name (\S\ref{sec:address}), disagree with them and argue (and can actually gain respect doing so), seek their help (even on personal matters), come to their houses for parties or gathering, and give them small thoughtful gifts that they put on their desks.  Many students and professors maintain lifelong relationships, staying in touch through cards, emails, and calls, even after their academic journey ends.% These connections are genuine and meaningful.

@@

\paragraph{About Me} I am an assistant professor in the CS dept at George Mason University (GMU). Prior to GMU, I was at the University of Nebraska-Lincoln (UNL). I have been in the PhD admission process at GMU and UNL for many years.  Currently I serve as the program director of the Masters program in Software Engineering at GMU (thus also have some experience with the MS admission process--which is quite different than PhD). I got my undergraduate and MS degrees from Penn State. I completed my PhD at the University of New Mexico-Albuquerque and a postdoc at the University of Maryland-College Park.

While I am not an international student, many of my students and collaborators are international.  I also mentor many students from Vietnam.  I also have many colleague and friends who were international students and have heard their stories.  I hope to capture their challenges and experiences in this document so that it can be useful for prospective international students.
